\documentclass[12pt,titlepage]{article}
\usepackage[a4paper,top=30mm,bottom=30mm,left=20mm,right=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{url}
\usepackage{alltt}
\usepackage{xspace}
\usepackage{times}
\usepackage{listings}
\usepackage{bbm}
\usepackage{verbatim}
\usepackage{optionman}

\newcommand{\LTRdigest}{\textit{LTRdigest}\xspace}
\newcommand{\GenomeTools}{\textit{GenomeTools}\xspace}
\newcommand{\Suffixerator}{\textit{Suffixerator}\xspace}
\newcommand{\GtLTRdigest}{\texttt{gt ltrdigest}\xspace}
\newcommand{\Gt}{\texttt{gt}\xspace}
\newcommand{\Gtsuffixerator}{\texttt{gt suffixerator}\xspace}

\title{\LTRdigest User's Manual}
\author{\begin{tabular}{c}
         \textit{Sascha Steinbiss}\\
         \textit{Ute Willhoeft}\\
         \textit{Gordon Gremme}\\
         \textit{Stefan Kurtz}\\[1cm]
         Research Group for Genome Informatics\\
         Center for Bioinformatics\\
         University of Hamburg\\
         Bundesstrasse 43\\
         20146 Hamburg\\
         Germany\\[1cm]
         \url{steinbiss@zbh.uni-hamburg.de}\\[1cm]
         \begin{tabular}{p{0.8\textwidth}}
        In any documentation or publication about research using \LTRdigest
        please cite the following paper:\\[5mm]
        S.~Steinbiss, U.~Willhoeft, G.~Gremme and S.~Kurtz.
        Fine-grained annotation and classification of \emph{de novo} predicted LTR
        retrotransposons.
        \emph{Nucleic Acids Research} 2009, 37(21):7002--7013\\[1mm]
        \url{http://nar.oxfordjournals.org/cgi/content/full/37/21/7002}
        \end{tabular}
        \end{tabular}}

\begin{document}
\maketitle

\section{Introduction}
\label{Introduction}

This document describes \LTRdigest , a software tool for identification and
annotation of characteristic sequence features of LTR retrotransposons in
predicted candidates, like those reported by\emph{LTRharvest}~\cite{EKW07}. In
particular, \LTRdigest can be used to find
\begin{itemize}
  \item protein domains,
  \item polypurine tracts (PPT) and
  \item primer binding sites (PBS)
\end{itemize}
inside a sequence region predicted to be a LTR retrotransposon.

For this identification, \LTRdigest utilises a number of algorithms to create an
annotation based on user-supplied constraints. For example, length and position
values for possible features can be extensively parameterised, as can be
algorithmic parameters like alignment scores and cut-off values.

\LTRdigest computes the boundaries and attributes of the features that fit the
user-supplied model and outputs them in GFF3 format~\cite{gff3} (in addition to
the existing LTR retrotransposon annotation), as well as the corresponding
sequences in multiple FASTA format. In addition, a tab-separated summary file is
created that can conveniently and quickly browsed for results.

\LTRdigest is written in C and it is based on the \GenomeTools
library~\cite{genometools}. It is called as part of the single binary named \Gt.

The source code can be compiled on 32-bit and 64-bit platforms without making
any changes to the sources. It incorporates HMMER~\cite{hmmer}, a popular and
widely used profile hidden Markov model package that is used for identification
of protein domains, for example using pHMMs taken from the Pfam~\cite{pfam}
database. The protein domain search is implemented to be run in a multi-threaded
fashion, thus making use of modern multi-core computer systems.

\section{Building \emph{LTRdigest}} \label{Building}

As \LTRdigest is part of the \GenomeTools software suite, a source distribution
of \GenomeTools must be obtained, e.g.\@ via the \GenomeTools home page
(\url{http://genometools.org}), and decompressed into a source directory:

\begin{verbatim}
$ tar -xzvf genometools-X.X.X.tar.gz
$ cd genometools-X.X.X
\end{verbatim}

Then, it suffices to call \texttt{make} to compile the source. To enable protein
domain search functionality, please also append the \texttt{with-hmmer=yes}
option to the \texttt{make} call. This option will make sure that a HMMER source
package is downloaded and compiled along with the \Gt binary. Note that the
\texttt{wget} executable must be available in the current PATH to do so
(alternatively, you can download HMMER 2.3.2 manually from
\texttt{ftp://selab.janelia.org/pub/software/hmmer/CURRENT/} and untar it in the
\texttt{src/external/} subdirectory).

\begin{verbatim}
$ make with-hmmer=yes
\end{verbatim}%$

If the \texttt{with-hmmer} option is not specified or set to \texttt{no},
protein domain finding functionality will be missing from the \LTRdigest
program. If the build process reports an error due to an unavailable Cairo
library, also append the \texttt{cairo=no} option to build the \Gt binary
without Cairo support. This has no influence on the function of \LTRdigest .

To enable multithreading support (that is, to speed up protein domain search by
searching for multiple pHMMs at once), please also specify the
\texttt{threads=yes} option to the \texttt{make} call:

\begin{verbatim}
$ make with-hmmer=yes threads=yes
\end{verbatim}%$

After successful compilation, the \GenomeTools executable containing \LTRdigest
can then be installed for system-wide use as follows:

\begin{verbatim}
$ make install
\end{verbatim}%$

If a \texttt{prefix=<path>} option is appended to this line, a custom directory
can be specified as the installation target directory, e.g.\@

\begin{verbatim}
$ make install prefix=/home/user/gt
\end{verbatim}%$
%$

will install the \Gt binary in the \texttt{/home/user/gt/bin} directory.
Please also consult the \texttt{README} and \texttt{INSTALL} files in the root
directory of the uncompressed source tree for more information and
troubleshooting advice.

\section{Usage} \label{Usage}

Some text is highlighted by different fonts according to the following rules.

\begin{itemize}
\item \texttt{Typewriter font} is used for the names of software tools.
\item \texttt{\small{Small typewriter font}} is used for file names.
\item \begin{footnotesize}\texttt{Footnote sized typewriter font}
      \end{footnotesize} with a leading
      \begin{footnotesize}\texttt{'-'}\end{footnotesize}
      is used for program options.
\item \Showoptionarg{small italic font} is used for the argument(s) of an
      option.
\end{itemize}

\subsection{\LTRdigest command line options}

Since \LTRdigest is part of \Gt, \LTRdigest is called as follows.

\GtLTRdigest  $[$\emph{options}$]$ \Showoptionarg{GFF3\_file} \Showoptionarg{indexname}

where \emph{GFF3\_file} denotes the GFF3 input file and \emph{indexname} the
name of an encoded sequence as created by the \Gtsuffixerator tool.  An overview
of all possible options with a short one-line description of each option is
given in Table \ref{overviewOpt}. They can also be displayed when invoking
\LTRdigest with the option \Showoption{help} or \Showoption{help+}.  All options
can be specified only once.

To run the protein domain search in a parallel fashion, use the \Showoption{j}
parameter to \texttt{gt} to specify the number of concurrent threads to use:

\texttt{gt}  \Showoption{j} 3 \texttt{ltrdigest} $[$\emph{options}$]$
\Showoptionarg{GFF3\_file} \Showoptionarg{indexname}

\begin{table}[htbp]
\caption{Overview of the \LTRdigest options sorted by categories.}
\begin{footnotesize}
\[
\renewcommand{\arraystretch}{0.89}
\begin{tabular}{ll}\hline
\Showoptiongroup{Input options}
\emph{GFF3\_file}& specify the path to the GFF3 input file
\\
\emph{indexname}& specify the path to the input sequences
\\

\Showoptiongroup{Output options}
\Showoption{outfileprefix}& specify prefix for sequence and tabular output files
\\
\Showoption{o}& specify file to output result GFF3 into
\\
\Showoption{gzip}& gzip-compress GFF3 output file specified by \Showoption{o}
\\
\Showoption{bzip2}& bzip2-compress GFF3 output file specified by \Showoption{o}
\\
\Showoption{force}& force output file to be overwritten
\\
\Showoption{aaout}& output amino acid sequences for protein domain hits
\\
\Showoption{aliout}& output HMMER amino acid alignments
\\
\Showoption{seqnamelen}& set maximal length of sequence names in output FASTA headers\\
                       & (e.g.\@ for clustalw or similar tools)
\\
\Showoptiongroup{PPT options}
\Showoption{pptlen}& specify a range of acceptable PPT lengths
\\
\Showoption{uboxlen}& specify a range of acceptable U-box lengths
\\
\Showoption{pptradius}& specify region around 3' LTR beginning to search for PPT
\\
\Showoption{pptrprob}& purine emission probability inside PPT
\\
\Showoption{pptyprob}& pyrimidine emission probability inside PPT
\\
\Showoption{pptaprob}& background A emission probability outside PPT
\\
\Showoption{pptcprob}& background C emission probability outside PPT
\\
\Showoption{pptgprob}& background G emission probability outside PPT
\\
\Showoption{ppttprob}& background T emission probability outside PPT
\\
\Showoption{pptuprob}& U/T emission probability inside U-box
\\
\Showoptiongroup{PBS options}
\Showoption{trnas}& tRNA library in multiple FASTA format
\\
\Showoption{pbsalilen}& specify a range of acceptable PBS lengths
\\
\Showoption{pbsoffset}& specify a range of acceptable PBS start distances from
3' end of 5' LTR
\\
\Showoption{pbstrnaoffset}& specify a range of acceptable tRNA/PBS alignment
offsets from tRNA 3' end
\\
\Showoption{pbsmaxedist}& specify the maximal allowed unit edit distance in
tRNA/PBS alignment
\\
\Showoption{pbsradius}& specify region around 5' LTR end to search for PBS
\\
\Showoptiongroup{Protein domain search options}
\Showoption{hmms}& specify a list of pHMMs for domain search in HMMER2 format
\\
\Showoption{pdomevalcutoff}& specify an E-value cutoff for pHMM search
\\
\Showoption{maxgaplen}& maximum allowed chaining gap size between fragments (in
amino acids)

\\
\Showoptiongroup{Alignment options}
\Showoption{pbsmatchscore}& specify matchscore for PBS/tRNA Smith-Waterman
alignment
\\
\Showoption{pbsmismatchscore}& specify mismatchscore for PBS/tRNA Smith-Waterman
alignment
\\
\Showoption{pbsinsertionscore}& specify insertionscore for PBS/tRNA
Smith-Waterman alignment
\\
\Showoption{pbsdeletionscore}& specify deletionscore for PBS/tRNA Smith-Waterman
alignment
\\
\Showoptiongroup{Miscellaneous options}
\Showoption{v}& verbose mode
\\
\Showoption{help}& show basic options
\\
\Showoption{help+}& show basic and extended options
\\
\hline
\end{tabular}
\]
\end{footnotesize}
\label{overviewOpt}
\end{table}

\newpage
%%%%
\subsection{Input parameters}

\emph{GFF3\_file}\\ specifies the path to the GFF3 input file. It has to include
at least position annotation for the LTR retrotransposon itself
(\texttt{LTR\_retrotransposon} type) and the predicted LTRs
(\texttt{long\_terminal\_repeat} type) because this information is needed to
locate the favored positions of the features in question. The GFF3 file must
also be sorted by position, which can be done using \GenomeTools :\\
\texttt{gt gff3} \Showoption{sort} \Showoptionarg{unsorted\_gff3\_file}
\texttt{\symbol{62}} \Showoptionarg{sorted\_gff3\_file}\\

\emph{indexname}\\ specifies the index name of an encoded sequence file
containing the sequences the GFF3 coordinates refer to. For each sequence in the
encoded sequence file given as the second parameter, there must exist a sequence
region in the GFF3 file named `\texttt{seq}\textit{i}' where $i$ is the
(zero-based) index number of the corresponding sequence in the encoded sequence.
For instance, all GFF3 feature coordinates for features on the first sequence in
the index file must be on sequence region \texttt{seq0}, and so on. An encoded
sequence can be created from a FASTA, GenBank or EMBL format file using the
\Gtsuffixerator command:\\
\Gtsuffixerator \Showoption{tis} \Showoption{des} \Showoption{dna}
\Showoption{ssp} \Showoption{db} \Showoptionarg{sequencefile}
\Showoption{indexname} \Showoptionarg{indexname}

%%%%
\subsection{Output options}

Results are reported in GFF3 format on stdout and can easily
be written to a file using the notation \texttt{\symbol{62}}
\Showoptionarg{GFF3\_resultfile} as in the following example:

\GtLTRdigest $[$\emph{options}$]$ \Showoptionarg{GFF3\_file}
\Showoptionarg{indexname} \texttt{\symbol{62}} \Showoptionarg{GFF3\_resultfile}

\begin{Justshowoptions}
\Option{outfileprefix}{\Showoptionarg{prefix}}{
If this option is given, a number of files containing further information will
be created during the \LTRdigest\ run:
\begin{itemize}
  \item \texttt{$<$prefix$>$\_tabout.csv} contains a tab-separated summary of
    the results that can, for example, be opened in a spreadsheet software or
    processed by a script. Each column is described in the file's header line
    and each row describes exactly one LTR retrotransposon candidate.
  \item \texttt{$<$prefix$>$\_conditions.csv} contains information about the
    parameters used in the current run for documentation purposes.
  \item \texttt{$<$prefix$>$\_pbs.fas} contains the PBS sequences identified in
    the current run in multiple FASTA format.
  \item \texttt{$<$prefix$>$\_ppt.fas} contains the PPT sequences identified in
    the current run in multiple FASTA format.
  \item The files \texttt{$<$prefix$>$\_5ltr.fas} and
    \texttt{$<$prefix$>$\_3ltr.fas} contain the 5' and 3' LTR sequences
    identified in the current run in multiple FASTA format. Please note: If the
    direction of the retrotransposon could be predicted, the files will contain
    the corresponding 3' and 5' LTR sequences. If no direction could be
    predicted, forward direction with regard to the original sequence will be
    assumed, i.e. the 'left' LTR will be considered the 5' LTR.
   \item Additionally, one \texttt{$<$prefix$>$\_pdom\_$<$domainname$>$.fas}
     file will be created per protein domain model given. This file contains the
     FASTA DNA sequences of the HMM matches to the LTR retrotransposon
     candidates.
\end{itemize}
In FASTA output files, each FASTA header contains position and sequence region
information to match the hit to the corresponding LTR retrotransposon.
}
\Option{aaout}{\Showoptionarg{yes/no}}{
If this option is set to \Showoptionarg{yes}, one
\texttt{$<$prefix$>$\_pdom\_$<$domainname$>$\_aa.fas} file will be created per
protein domain model given. This file contains the (concatenated) FASTA amino
acid sequences of the HMM matches to the LTR retrotransposon candidates.
}
\Option{aliout}{\Showoptionarg{yes/no}}{
If this option is set to \Showoptionarg{yes}, one
\texttt{$<$prefix$>$\_pdom\_$<$domainname$>$.ali} file will be created per
protein domain model given. This file contains alignment information for all
matches to of the given protein domain model to the translations of all
candidate.
}
\end{Justshowoptions}


%%%%
\subsection{PPT options}

\begin{Justshowoptions}

\Option{pptlen}{\Showoptionarg{$L_{min}$} \Showoptionarg{$L_{max}$}}{
Specify the minimum and maximum allowed lengths for PPT predictions. If a
purine-rich region shorter than \Showoptionarg{$L_{min}$} or longer than
\Showoptionarg{$L_{max}$} is found, it will be skipped.
\\
\Showoptionarg{$L_{min}$} and \Showoptionarg{$L_{min}$} have to be positive
integers. If this option is not selected, then \Showoptionarg{$L_{min}$} is set
to $8$, \Showoptionarg{$L_{max}$} to 30.
}

\Option{uboxlen}{\Showoptionarg{$L_{min}$} \Showoptionarg{$L_{max}$}}{
Specify the minimum and maximum allowed lengths for U-box predictions. If a
T-rich region preceding a PPT shorter than \Showoptionarg{$L_{min}$} or longer
than \Showoptionarg{$L_{max}$} is found, it will be skipped.\\
\Showoptionarg{$L_{min}$} and \Showoptionarg{$L_{min}$} have to be positive
integers. If this option is not selected, then \Showoptionarg{$L_{min}$} is set
to $3$, \Showoptionarg{$L_{max}$} to 30.
}

\Option{pptradius}{\Showoptionarg{$r$}}{
Specify the area around the 3' LTR beginning ($l_{s}$) to be searched for PPTs,
in other words, define the search interval $[l_{s}-r, l_{s}+r]$.\\
\Showoptionarg{$r$} has to be a positive integer. If this option is not
selected, then \Showoptionarg{$r$} is set to 30.
}

\Option{pptrprob}{\Showoptionarg{$p_R$}}{
  Specify the emission probability of a purine base (\texttt{A}/\texttt{G})
  inside a PPT.  This value must be a valid probability value ($0\leq p_R \leq
  1$).  If this option is not set, then \Showoptionarg{$p_R$} is set to 0.97.
}
\Option{pptyprob}{\Showoptionarg{$p_Y$}}{
  Specify the emission probability of a pyrimidine base (\texttt{T}/\texttt{C})
  inside a PPT.  This value must be a valid probability value ($0\leq p_Y \leq
  1$).  If this option is not set, then \Showoptionarg{$p_Y$} is set to 0.03.
}
\Option{pptaprob}{\Showoptionarg{$p_A$}}{
  Specify the background emission probability of an \texttt{A} base outside of
  PPT and U-box regions.  This value must be a valid probability value ($0\leq
  p_A \leq 1$).  If this option is not set, then \Showoptionarg{$p_A$} is set to
  0.25.
}
\Option{pptcprob}{\Showoptionarg{$p_C$}}{
  Specify the background emission probability of a \texttt{C} base outside of
  PPT and U-box regions.  This value must be a valid probability value ($0\leq
  p_C \leq 1$).  If this option is not set, then \Showoptionarg{$p_C$} is set to
  0.25.
}
\Option{pptgprob}{\Showoptionarg{$p_G$}}{
  Specify the background emission probability of a \texttt{G} base outside of
  PPT and U-box regions.  This value must be a valid probability value ($0\leq
  p_G \leq 1$).  If this option is not set, then \Showoptionarg{$p_G$} is set to
  0.25.
}
\Option{ppttprob}{\Showoptionarg{$p_T$}}{
  Specify the background emission probability of a \texttt{T} base outside of
  PPT and U-box regions.  This value must be a valid probability value ($0\leq
  p_T \leq 1$).  If this option is not set, then \Showoptionarg{$p_T$} is set to
  0.25.
}
\end{Justshowoptions}
Note that $\sum_{x \in \left\{A,C,G,T\right\}} p_x = 1$ must hold if the
\Showoption{ppt\{a,c,g,t\}prob} options are used.
\begin{Justshowoptions}


\Option{pptuprob}{\Showoptionarg{$p_U$}}{
  Specify the emission probability of a thymine base (\texttt{T}) inside a
  U-box.  This value must be a valid probability value ($0<p_U \leq 1$). All
  other emission probabilities are calculated as uniform probabilities $p_x =
  \frac{1-p_U}{3}$ for all $x \in \left\{A,C,G\right\}$.  If this option is not
  set, then \Showoptionarg{$p_U$} is set to 0.91.
}

\end{Justshowoptions}

\subsection{PBS options}

\begin{Justshowoptions}

\Option{trnas}{\Showoptionarg{$trnafile$}}{
Specify a file in multiple FASTA format to be used as a tRNA library that is
aligned to the area around the end of the 5' LTR to find a putative PBS. The
header of each sequence in this file should reflect the encoded amino acid and
codon.  If this option is not selected, then PBS searching is skipped
altogether.
}

\Option{pbsalilen}{\Showoptionarg{$L_{min}$} \Showoptionarg{$L_{max}$}}{
Specify the minimum and maximum allowed lengths for PBS/tRNA alignments. If a
local alignment shorter than \Showoptionarg{$L_{min}$} or longer than
\Showoptionarg{$L_{max}$} is found, it will be skipped.  \\
\Showoptionarg{$L_{min}$} and \Showoptionarg{$L_{min}$} have to be positive
integers. If this option is not selected, then \Showoptionarg{$L_{min}$} is set
to $11$, \Showoptionarg{$L_{max}$} to 30.
}

\Option{pbsoffset}{\Showoptionarg{$L_{min}$} \Showoptionarg{$L_{max}$}}{ Specify
  the minimum and maximum allowed distance between the start of the PBS and the
  3' end of the 5' LTR. If a local alignment with such a distance smaller than
  \Showoptionarg{$L_{min}$} or greater than \Showoptionarg{$L_{max}$} is found,
  it will be skipped.  \\
\Showoptionarg{$L_{min}$} and \Showoptionarg{$L_{min}$} have to be positive
integers. If this option is not selected, then \Showoptionarg{$L_{min}$} is set
to $0$, \Showoptionarg{$L_{max}$} to 5.
}

\Option{pbstrnaoffset}{\Showoptionarg{$L_{min}$} \Showoptionarg{$L_{max}$}}{
Specify the minimum and maximum allowed PBS/tRNA alignment offsets from the 3'
end of the tRNA. If a local alignment with an offset smaller than
\Showoptionarg{$L_{min}$} or greater than \Showoptionarg{$L_{max}$} is found, it
will be skipped.  \\
\Showoptionarg{$L_{min}$} and \Showoptionarg{$L_{min}$} have to be positive
integers. If this option is not selected, then \Showoptionarg{$L_{min}$} is set
to $0$, \Showoptionarg{$L_{max}$} to 5.
}

\Option{pptmaxedist}{\Showoptionarg{$d$}}{
Specify the maximal allowed unit edit distance in a local PBS/tRNA alignment.
All optimal local alignments with a unit edit distance $> d$ will be skipped.
Set this to 0 to accept exact matches only. It is also possible to fine-tune the
results by adjusting the match/mismatch/indelscores used in the Smith-Waterman
alignment (see below).\\
\Showoptionarg{$d$} has to be a positive integer. If this option is not
selected, then \Showoptionarg{$d$} is set to 1.
}

\Option{pbsradius}{\Showoptionarg{$r$}}{
Specify the area around the 5' LTR end ($l_{e}$) to be searched for a PBS, in
other words, define the search interval $[l_{e}-r, l_{e}+r]$.\\
\Showoptionarg{$r$} has to be a positive integer. If this option is not
selected, then \Showoptionarg{$r$} is set to 30.
}

\Option{pbsmatchscore}{\Showoptionarg{$score_m$}}{
Specify the match score used in the PBS/tRNA Smith-Waterman alignment. Lower
this value to discourage matches, increase this value to prefer matches.\\
\Showoptionarg{$score_m$} has to be an integer. If this option is not selected,
then \Showoptionarg{$score_m$} is set to 5.
}

\Option{pbsmismatchscore}{\Showoptionarg{$score_{mm}$}}{
Specify the mismatch score used in the PBS/tRNA Smith-Waterman alignment. Lower
this value to discourage mismatches, increase this value to prefer mismatches.\\
\Showoptionarg{$score_{mm}$} has to be an integer. If this option is not
selected, then \Showoptionarg{$score_{mm}$} is set to -10.
}

\Option{pbsdeletionscore}{\Showoptionarg{$score_d$}}{
Specify the deletion score used in the PBS/tRNA Smith-Waterman alignment. Lower
this value to discourage deletions, increase this value to prefer deletions.\\
\Showoptionarg{$score_d$} has to be an integer. If this option is not selected,
then \Showoptionarg{$score_d$} is set to -20.
}

\Option{pbsinsertionscore}{\Showoptionarg{$score_i$}}{
Specify the insertion score used in the PBS/tRNA Smith-Waterman alignment. Lower
this value to discourage insertions, increase this value to prefer insertions.\\
\Showoptionarg{$score_i$} has to be an integer. If this option is not selected,
then \Showoptionarg{$score_i$} is set to -20.
}

\end{Justshowoptions}

\subsection{Protein domain search options}

\begin{Justshowoptions}

\Option{hmms}{\Showoptionarg{$hmmfile_1, hmmfile_2, \dots, hmmfile_n$}}{
Specify a list of pHMM files in HMMER2 format. The pHMMs must be defined for the
amino acid alphabet and follow the Plan7 specification. For example, pHMMs
defining protein domains taken from the Pfam database can be used here. Every
file must exist and be readable, otherwise an error is reported. If this option
is not given, protein domain searching is skipped altogether. Please note that
shell globbing can be used here to specify large numbers of files, e.g.\@ by using
wildcards.}

\Option{pdomevalcutoff}{\Showoptionarg{$c$}}{
Specify the E-value cutoff for HMMER searches. All hits that fail to meet this
maximal e-value requirement are discarded.\\
\Showoptionarg{$c$} has to be a probability ($0 \leq c \leq 1$). If this option
is not selected, then \Showoptionarg{$c$} is set to $10^{-6}$.
}

\end{Justshowoptions}

\section{Example}

This section describes an example session with \LTRdigest . For simplicity, we
assume that a \emph{LTRharvest} run on a genome has already been performed and
produced a \texttt{ltrs.gff3} file containing the basic GFF3 annotation
describing LTR positions. The enhanced suffix array (ESA) index, any sequence
output or the tabular standard output from \emph{LTRharvest} will not be needed.
Instead, we require an encoded sequence representation of the original input
sequence, in this example called \texttt{genome.fas}. This can be created using
the \GenomeTools \Gtsuffixerator tool as follows:

\texttt{gt suffixerator -tis -des -dna -ssp -db genome.fas -indexname
genome.fas}

This step creates additional files with the suffixes \texttt{.al1},
\texttt{.des}, \texttt{.esq}, \texttt{.prj} and \texttt{.ssp} in the sequence
file directory.

We will also assume that the HMM files to be used are called \texttt{HMM1.hmm},
\texttt{HMM2.hmm} and \texttt{HMM3.hmm}, we will be using a tRNA library called
\texttt{tRNA.fas}, and we are running \LTRdigest on a dual-core system. We also
want to restrict the PBS offset from the LTR end to a maximum of 3 nucleotides,
while we want the PPT length to be at least 10 nucleotides. Finally, we want all
output such as sequences written to files beginning with ``mygenome-ltrs''.

First, sort the GFF3 output by position:
\\[0.5cm]
\texttt{gt gff3 -sort ltrs.gff3 $>$ ltrs\_sorted.gff3}
\\[0.5cm]
Then, the \LTRdigest run can be started with the following command line using
the parameters above:
\\[0.5cm]
\texttt{gt -j 2 ltrdigest -pptlen 10 30 -pbsoffset 0 3 -trnas tRNA.fas \\-hmms
HMM*.hmm -outfileprefix mygenome-ltrs ltrs\_sorted.gff3 genome.fas $>$
ltrs\_after\_ltrdigest.gff3}

No screen output (except possible error messages) is produced since the GFF3
output on stdout is redirected to a file. Additionally, the files
\texttt{mygenome\--ltrs\_conditions.csv}, \texttt{mygenome\--ltrs\_3ltr.fas},
\texttt{mygenome\--ltrs\_5ltr.fas}, \texttt{mygenome\--ltrs\_ppt.fas},
\texttt{mygenome\--ltrs\_pbs.fas}, \texttt{mygenome-\-ltrs\_tabout.csv} and one
FASTA file for each of the HMM models will be created and updated during the
computation. As the files are buffered, it may take a while before first output
to these files can be observed.

The calculation may be restarted with the same or changed parameters afterwards,
overwriting the output files in the process. If it is desired to keep sequences
etc. from each run, keep in mind to assign specific \texttt{-outfileprefix}
values to each run.

\bibliographystyle{unsrt}
\begin{thebibliography}{1}

\bibitem{EKW07}
D.~Ellinghaus, S.~Kurtz, and U.~Willhoeft.
\newblock \emph{LTRharvest}, an efficient and flexible software for de novo
detection of \normalsize{LTR} retrotransposons.
\newblock {\em BMC Bioinformatics} 9:18, 2008.

\bibitem{gff3}
L.~Stein.
\newblock Generic Feature Format Version 3.
  \url{http://www.sequenceontology.org/gff3.shtml}.

\bibitem{genometools}
G.~Gremme.
\newblock GenomeTools.
  \url{http://genometools.org}.

\bibitem{pfam}
R.D.~Finn, J.~Mistry, B.~Schuster-Boeckler, S.~Griffiths-Jones, V.~Hollich,
T.~Lassmann,S.~Moxon, M.~Marshall, A.~Khanna, R.~Durbin, S.R.~Eddy,
E.L.L.~Sonnhammer and A.~Bateman.
\newblock Pfam: clans, web tools and services.
\newblock {\em Nucleic Acids Research} (Database Issue), 34:D247-D251, 2006.

\bibitem{hmmer}
S.R.~Eddy.
\newblock HMMER: Biosequence analysis using profile hidden Markov models.
  \url{http://hmmer.janelia.org}.

\end{thebibliography}
\end{document}
