\documentclass[12pt,titlepage]{article}
\usepackage[a4paper,top=30mm,bottom=30mm,left=20mm,right=20mm]{geometry}
\usepackage{url}
\usepackage{alltt}
\usepackage{xspace}
\usepackage{times}
\usepackage{listings}
\usepackage{bbm}
\usepackage{verbatim}
\usepackage{comment}
\usepackage{optionman}

\newcommand{\packedindex}{\textit{packedindex}\xspace}
\newcommand{\GenomeTools}{\textit{GenomeTools}\xspace}
\newcommand{\Suffixerator}{\textit{Suffixerator}\xspace}
\newcommand{\Gt}{\texttt{gt}\xspace}
\newcommand{\Gtsuffixerator}{\texttt{gt suffixerator}\xspace}
\newcommand{\Gtpackedindex}{\texttt{gt packedindex}\xspace}
\newcommand{\species}[1]{{\fontshape{it}\selectfont #1}}%

\title{\packedindex  manual}
\author{\begin{tabular}{c}
         \textit{Thomas Jahns}\\
         \textit{Stefan Kurtz}\\[1cm]
         Research Group for Genomeinformatics\\
         Center for Bioinformatics\\
         University of Hamburg\\
         Bundesstrasse 43\\
         20146 Hamburg\\
         Germany\\[1cm]
         \url{kurtz@zbh.uni-hamburg.de}\\[1cm]
         \begin{tabular}{p{0.8\textwidth}}
%        In any documentation or publication about research using \LTRharvest
%        please cite the following paper:\\[5mm]
%        D.~Ellinghaus, S.~Kurtz, and U.~Willhoeft.
%        \LTRharvest, an efficient and flexible software for de novo
%        detection of LTR retrotransposons.
%        BMC Bioinformatics 2008, 9:18
        \end{tabular}
        \end{tabular}}
\begin{document}
\maketitle

\section{Introduction}
\label{Introduction}

The \packedindex tools of the \GenomeTools package facilitate the
creation of sequence index files which implement the FM
index\cite{oai:CiteSeerPSU:481302} concept for biological sequences
with some more recent features
incorporated\cite{journals/talg/FerraginaMMN07}.

The \packedindex routines and tools are written in \texttt{C} and are
part of the \GenomeTools library \cite{genometools}. \packedindex
methods are called as part of the single binary named \Gt.

The \packedindex tools depend on routines of the \Suffixerator program
to prepare the  data stored in the index. Many command-line options
are therefore the same.

\section{Command line tools}
\label{Usage}

Some text is highlighted by different fonts according to the following rules.

\newcommand{\toolname}[1]{\texttt{\bfseries #1}}
\newcommand{\toolarg}[1]{\texttt{\itshape #1}}
\newcommand{\tooloption}[1]{\texttt{-#1}}
\newcommand{\tooloptionarg}[1]{\texttt{#1}}
\begin{itemize}
\item \toolname{toolname} is used for the names of software tools.
\item \toolarg{filename} is used for file names.
\item \tooloption{option} is used for program options.
\item \tooloptionarg{argument} is used for the argument(s) of an
      option.
\end{itemize}

\subsection{The \packedindex tools}
\label{Overview}

The creation and integrity checks of the \packedindex tools are
available as a sub-tool of \Gt. Each of those is called by giving
\toolname{packedindex} as the first argument of the \toolname{gt}
command-line.

%\Gtltrharvest \Showoption{index} \Showoptionarg{indexname} $[$\emph{options}$]$%
%
%where \Showoptionarg{indexname} denotes the enhanced suffix array index of 
%the target sequence(s) constructed by the program \Suffixerator. 
%An overview of all possible options with a short one-line description of 
%each option is given in Table \ref{overviewOpt}.
%All options can be specified only once.

\subsubsection{Options common to all tools}
\label{sec:packedindex:commonoptions}

\begin{Justshowoptions}

% \Option{v}{}{
% This option enables the verbose mode. This means, that some more information
% about the processing will be printed to \texttt{stdout} during the run.
% This includes a long list of switched on or switched off options.
% }

\Option{help}{}{
Prints a summary of available options to standard output and
terminates with exit code $0$.
}

\Option{v}{}{
Turns on verbose progress information.
}

\Option{version}{}{
Displays \toolname{genometools} version information.
}

\end{Justshowoptions}


\subsubsection{The creation of packedindex files}
\label{sec:packedindex:mkindex}

The command to build a \packedindex is
\begin{quote}
  \ttfamily%
  \toolname{gt} \toolname{packedindex} \toolname{mkindex} [{\rmfamily
    Options}] \tooloption{db} \toolarg{sequenceinput...}
\end{quote}

The following options are specific to the \packedindex
\toolname{mkindex} tool:

\begin{table}[htbp]
\caption{Overview of the \packedindex \toolname{mkindex} options.}
\begin{footnotesize}
\[
\renewcommand{\arraystretch}{0.89}
\begin{tabular}{ll}\hline
\Showoption{bsize}& specify the number of symbols joined in a block
\\
\Showoption{blbuck}& specify number of blocks to join in a bucket
\\
\Showoption{locfreq}& specify interval at which locate marks will be inserted
\\
\Showoption{locbitmap}& force/deny use of the bitmap tagging method
\\
\Showoption{sprank}& build source rank table so that regeneration of
original sequence is possible without the usual gaps
\\
\Showoption{sprankilog}& define log of interval used for the
intermediate table
\\
\hline
\end{tabular}
%\input{ltrprogoptions}
\]
\end{footnotesize}
\label{tab:packedindex:mkindex:options}
\end{table}

Except for the output options \Showoption{suf}, \Showoption{lcp}, and
\Showoption{bwt}, all other options are the same as used for the
\toolname{suffixerator} tool.

%%%%
\begin{Justshowoptions}

\Option{bsize}{\Showoptionarg{u}}{
Group \Showoptionarg{u} symbols in one block. While bigger blocks can
speed up index access, the block size should be chosen so that
\begin{displaymath}
  \alpha^u u\lceil\log_2 \alpha\rceil
\end{displaymath}
does not exceed the relevant cache size (typically of the 2nd level
chache) of the system.

The default of a block size of 8 results in a lookup-table size of
128kiB for the DNA alphabet. Systems with larger 2nd level cache like
4MiB on the Intel Core2 Duo might show improved performance when
choosing block size 10.
}

\Option{blbuck}{\Showoptionarg{k}}{
Group this many blocks so that every atomic query of the index will on
average require $\frac{k}{2}$ lookups of the block table (see option
\Showoption{bsize}). While smaller blocks decrease query times, the
corresponding increase in memory usage of the index will eventually
defeat this effect.
}

\Option{locfreq}{\Showoptionarg{f}}{
Put location marks at interval \Showoptionarg{f} into the
index. For a sequence of length $n$, $\frac{n}{f}\lceil\log_2n\rceil$ bits of
storage will be required. 

Locate queries (i.e. finding the position of matches) will
require $\frac{f}{2}$ operations. Again having an index that does not
fit main memory will eventually defeat the speed gains of lower
intervals.
}

\Option{locbitmap}{\Showoptionarg{yes} or \Showoptionarg{no}}{
Force/disable the use of a bitmap for locate marks. Finding marked
positions will be faster if a bitmap is used but will require $n$ bits
for a sequence of length $n$. If no locate bitmap is used, extra space usage
will be $\frac{n}{f}\lceil\log_2uk\rceil$.

By default the optimal value in terms of space usage will be chosen.
}
\Option{sprank}{\Showoptionarg{yes} or \Showoptionarg{no}}{
Normally special characters (N/X, separators and terminator) are
encoding so that the reverse mapping required for context regeneration
cannot pass over such symbols. If this option is given, an in-memory
structure will be built to facilitate queries which allow to
store necessary extra information in the index so that full context
(the entire original sequence actually) can be regenerated.
}
\Option{sprankilog}{\Showoptionarg{$s$}}{
Set logarithm of interval used for the intermediate table (see
\Showoption{sprank} option).
Index creation might be faster if a lower value
than the default $s = \log_2\log_2n$ is used and sufficient memory is
available.

Since every increment by 1 doubles the memory usage of the table
(which is not stored itself in the index) during index construction
two high values might actually slow index construction.
}

\Option{ctxilog}{\Showoptionarg{$c$}}{
Set logarithm of interval to use for the creation of a reverse lookup
table. Said table will be stored alongside the index in a file with
the same name prefix but suffix \toolarg{.$c$cxm} so that multiple
reverse lookup accelerator tables can be stored for one index. So it's
possible to load the one fitting available memory best later.

The reverse lookup table is restricted to indices computed with the
\Showoption{sprank} option. In those cases the index can replace the
encoded sequence (stored in \toolarg{prefix.esq}) completely.

The reverse suffix indices will be sampled at interval length $2^c$
}
\end{Justshowoptions}
%%%%

\subsubsection{Building reverse lookup tables after index creation}
\label{sec:packedindex:mkctxmap}

The \toolname{mkctxmap} sub-tool facilitates late building of a
reverse lookup table, it can use the \toolarg{prefix.suf} table as
well as the locate information inside the packedindex structure
(i.e. of any packedindex not constructed with \tooloption{locfreq} set
to 0).

The \toolname{mkctxmap} sub-tool only accepts the
\tooloption{-ctxilog}, \tooloption{-v}, \tooloption{-version}, and
\tooloption{-help} options already described for the
\toolname{packedindex mkindex} tool.

Given a previously computed suffix array or packedindex project
\toolarg{chr01}, one can build an additional reverse lookup-table with
interval 32 like this:
\begin{footnotesize}
\begin{verbatim}
$ gt packedindex mkctxmap -ctxilog 5 chr01
\end{verbatim}
\end{footnotesize}

\subsubsection{Deriving packedindex files from a suffix array}
\label{sec:packedindex:trsuftab}

The \toolname{trsuftab} sub-tool will translate an index by reading
suffix array and encoded sequence files. Thus the options specific to
the packedindex properties are the same as previously
shown~\ref{sec:packedindex:mkindex} but the suffixerator options are
not available. Instead the latter are implicit in the properties of
the suffix array and encoded sequence indices used as input for this
tool.

Thus to produce a packedindex with default options where a suffix
array has already been created in chr01.\{prj$|$suf$|$esq\} use:
\begin{footnotesize}
\begin{verbatim}
$ gt packedindex trsuftab chr01
\end{verbatim}
\end{footnotesize}

\subsubsection{packedindex integrity check tool}
\label{sec:packedindex:chkintegrity}

The integrity check tool is intended to verify that an index correctly
represents the encoded BWT sequence. Therefore the .bwt file for the
index project in question must have been generated as a prerequisite
for this check.

To check a \packedindex index file BWT sequence representation issue
\begin{quote}
  \ttfamily%
  \toolname{gt} \toolname{packedindex} \toolname{chkintegrity} [{\rmfamily
    Options}] \toolarg{indexname}
\end{quote}

where \toolarg{indexname} refers to the name given previously on index
creation (or automatically derived from the sequence input file set).
The .bwt table file must have previously been generated with the
\toolname{suffixerator} tool. See
section~\ref{sec:packedindex:examples} for an example.

\begin{table}[htbp]
\caption{Overview of the \packedindex \toolname{chkintegrity} options.}
\begin{footnotesize}
\[
\renewcommand{\arraystretch}{0.89}
\begin{tabular}{ll}\hline
\Showoption{skip}& specify the number of symbols to skip before
starting the comparison
\\
\Showoption{ticks}& print a dot after this many symbols processed correctly
\\
\Showoption{ext-rank-check}& do additional checks of rank query results
\\
\hline
\end{tabular}
%\input{ltrprogoptions}
\]
\end{footnotesize}
\label{tab:packedindex:chkintegrity:options}
\end{table}

\begin{Justshowoptions}

\Option{skip}{\Showoptionarg{k}}{
Start the comparison \Showoptionarg{k} has to be a positive integer. If this
option is not selected by the user, then \Showoptionarg{$L_{ex}$} is set
to $30$ by default.
}

\Option{ticks}{\Showoptionarg{$k$}}{ 
As a form of progress meter, print a dot after $k$ symbols of the
packedindex and the reference file (\toolarg{indexname.bwt}) compared
equal and rank information was computed correctly.
}

\Option{ext-rank-check}{\Showoptionarg{yes$\|$no}}{ 
If yes, enables additional checks of the rank results, i.e. for all
positions rank queries will be computed for all symbols of the
alphabet and not only for the symbol for which the rank value changed
in the reference.
}
\end{Justshowoptions}

\subsubsection{Check \packedindex search accuracy tool}

This check verifies that a packedindex correctly represents both the
original sequence and correctly delivers the exact same search results
as the suffix array.

To check a \packedindex index file issue
\begin{quote}
  \ttfamily%
  \toolname{gt} \toolname{packedindex} \toolname{chksearch} [{\rmfamily
    Options}] \toolarg{indexname}
\end{quote}

where \toolarg{indexname} refers to the name given previously on index
creation (or automatically derived from the sequence input file set).

The encoded sequence (\toolarg{.esq}) and suffix array table
(\toolarg{.suf}) must have been previously generated with the
\toolname{suffixerator} tool.

\begin{table}[htbp]
\caption{Overview of the \packedindex \toolname{chksearch} options.}
\begin{footnotesize}
\[
\renewcommand{\arraystretch}{0.89}
\begin{tabular}{ll}\hline
\Showoption{minpatlen}& shortest pattern length
\\
\Showoption{maxpatlen}& longest pattern length
\\
\Showoption{ticks}& print a dot after this many symbols processed correctly
\\
\Showoption{nsamples}& generate this many patterns to search for
\\
\Showoption{chksfxarray}& do additional checks of stored suffix array data
\\
\hline
\end{tabular}
%\input{ltrprogoptions}
\]
\end{footnotesize}
\label{tab:packedindex:chksearch:options}
\end{table}

\begin{Justshowoptions}

\Option{minpatlen}{\Showoptionarg{$l$}}{ 
Perform searches for patterns that are at least of length $l$.
}

\Option{maxpatlen}{\Showoptionarg{$m$}}{
Perform searches for patterns that are at most of length $l$.
}

\Option{nsamples}{\Showoptionarg{$k$}}{
Generate $k$ patterns to search for, defaults to 1000.
}

\Option{chksfxarray}{\Showoptionarg{yes$\|$no}}{
Verify integrity of stored suffix array positions and test the
regenerated context functionality if the index was constructed in a
reversible fashion (option \Showoption{sprank} on index construction).

In the latter case the test will take considerable time.
}

\Option{ticks}{\Showoptionarg{$k$}}{
Print a dot after this many searches completed correctly. If
\Showoption{chksfxarray} is set, also prints a dot for every $k$
suffix array values compared.
}
\end{Justshowoptions}

\section{Examples}
\label{sec:packedindex:examples}

\subsection{Create an index for sequence data from the
  \toolarg{gttestdata} repository\cite{gttestdata}}
\label{sec:packedindex:examples:yeast}

Amongst the more voluminous sequence data sets there is the
\species{Saccharomyces cerevisiae} genome in the
\toolarg{ltrharvest/s\_cer} sub-directory. This example will show how
to build an index for chromosome 1 (file
\toolarg{chr01.19960731.fsa.gz}).

First compute the \packedindex index file (this assumes
\toolarg{chr01.19960731.fsa.gz} has been copied/linked to the current
directory and the gt binary is in the PATH):
\begin{footnotesize}
\begin{verbatim}
$ gt packedindex mkindex -db chr01.19960731.fsa.gz -indexname chr01 -tis -des
# TIME determining sequence length and number of special symbols 0.04
# TIME computing sequence encoding 0.04
[...]
# TIME overall 0.71
$
\end{verbatim}
\end{footnotesize}

Then build the reference files that the index subsumes:
\begin{footnotesize}
\begin{verbatim}
$ gt suffixerator -db chr01.19960731.fsa.gz -indexname chr01 -bwt -suf -des
# TIME determining sequence length and number of special symbols 0.04
# TIME computing sequence encoding 0.04
[...]
# TIME overall 0.31
$
\end{verbatim}
\end{footnotesize}

This index can then be verified or used in one of the tools supporting
it. For integrity check use:
\begin{footnotesize}
\begin{verbatim}
$ gt packedindex chkintegrity chr01
# Using index over sequence 230210 symbols long.
[...]
..
$ gt packedindex chksearch chr01
Using pre-computed sequence index.
[...]
Finished 1000 of 1000 matchings successfully.
$
\end{verbatim}
\end{footnotesize}

\bibliographystyle{unsrt}
\bibliography{gtmanuals}

\end{document}
