\documentclass[12pt]{article}
\usepackage[a4paper,top=20mm,bottom=20mm,left=20mm,right=20mm]{geometry}
\usepackage{url}
\usepackage{alltt}
\usepackage{xspace}
\usepackage{times}
\usepackage{bbm}
\usepackage{verbatim}
\usepackage{ifthen}
\usepackage{optionman}

\newcommand{\Substring}[3]{#1[#2..#3]}

\newcommand{\Program}[0]{\texttt{tagerator}\xspace}

\title{\Program: a program for mapping short sequence tags\\
       a manual}

\author{\begin{tabular}{c}
         \textit{Stefan Kurtz}\\
         Center for Bioinformatics,\\
         University of Hamburg
        \end{tabular}}

\begin{document}
\maketitle

\section{Preliminary definitions}
By \(S\) let us denote the concatenation of all subject sequences.
By \(edist(u,v)\) we denote the unit edit distance  of two strings
\(u\) and \(v\).
Consider a sequence \(p\) of length \(m\). An approximate match of \(p\) with 
up to \(k\) differences is a substring \(v\) of \(S\) such that 
\(edist(p,v)\leq k\). An approximate prefix match of \(p\) with \(k\) 
differences and at most \(t\) occurrences is a substring \(v\) of \(S\) such 
that \(v\) occurs at most \(t\) times in \(S\) and 
\(edist(v,\Substring{p}{1}{i})\leq k\) for some \(i\in[1,m]\).

\section{The program \Program}

The program \Program is called as follows:
\par
\noindent\texttt{gt} \Program [\textit{options}] \Showoption{q} \Showoptionarg{files} [\textit{options}] 
\par
\Showoptionarg{files} is a white space separated list of at least one
filename. Any sequence occurring in any file specified in
\Showoptionarg{files} is called \textit{short sequence tag} or \textit{tag}
for short. In addition to the mandatory option \Showoption{q}, the program
must be called with either option \Showoption{pck} or \Showoption{esa},
which specify to use a packed index and an enhanced suffix array,
respectively. Both indices are constructed from a given set of subject 
sequences. 

\Program maps each short sequence tag, say \(p\) of length \(m\)
against the given index. The length of the tag is limited by the
size of a pointer. If \texttt{gt} is a 32-bit binary, then \(m\) must be
smaller or equal to 32. If \texttt{gt} is a 64-bit binary, \(m\) must be 
smaller or equal to 64. The program runs in three basic modes:
\begin{itemize}
\item
In the \textit{ms}-mode, it computes for all \(i\in[1,m]\) the length \(\ell\) 
of the longest prefix of \(\Substring{w}{i}{m}\) matching any substring of 
\(S\).
In addition, it reports start positions of such a prefix in \(S\).
As these values make up the well known matching statistics, we denote 
it by \textit{ms}. The length value \(\ell\) is the matching statistics length
and the position values are the matching statistics positions.
Both, the matching statistics length and the matching statistics position
is uniquely determined. The matching statistics mode requires to specify 
a maximum number of occurrences, see option \Showoption{maxocc}.
\item
In the \textit{cdiff}-mode, it computes all start positions of approximate 
matches with up to \(k\) differences in \(S\). For each start position of 
an approximate match, say \(j\), it reports the minimum integer \(\ell\) such 
that \(edist(p,\Substring{S}{j}{j+\ell-1})\leq k\). Since this mode matches
the complete sequence \(p\), we call this mode \textit{cdiff},
for complete difference.
\item
In the \textit{pdiff}-mode, it computes all start positions of approximate 
prefix matches with up to \(k\) differences and at most \(t\) occurrences in 
\(S\). For each start position of an approximate prefix match, say \(j\), it 
reports the minimum integers \(i\) and \(\ell\) such that 
\(edist(\Substring{p}{1}{i},\Substring{S}{j}{j+\ell-1})\leq k\). Since this 
mode matches a prefix of \(p\) with some differences, we call this mode 
\textit{pdiff}.  This mode requires to specify a maximum number of 
occurrences, see option \Showoption{maxocc}.
\end{itemize}

The following options are available in \Program:

\begin{Justshowoptions}
\Option{q}{$\Showoptionarg{files}$}{
Specify a white space separated list of query files (in multiple \Fasta format)
containing the tags. At least one query file must be given. The files may be in 
gzipped format, in which case they have to end with the suffix \texttt{.gz}.
}

\Option{esa}{$\Showoptionarg{indexname}$}{
Use the given enhanced suffix array index to map the short sequence tags.
}

\Option{pck}{$\Showoptionarg{indexname}$}{
Use the packed index (an efficient representation of the FMindex)
to map the short sequence tags.
}

\Option{e}{$\Showoptionarg{k}$}{
Specify the number of differences allowed. \(k\) must be a non-negative number.
\(k=0\) means that no differences are allowed (exact matching) and \(k>0\)
means a positive number of differences. If this option is not used, then
the program runs in \textit{ms}-mode, i.e.\ it computes the matching statistics
for each short sequence tag.
}

\Option{nod}{}{
Do not compute direct matches, i.e.\ matches on the forward strand. If
this option is not used, then matches are computed on the forward strand.}

\Option{nop}{}{
Do not compute palindromic matches, i.e.\ matches on the 
reverse complemented strand.  If
this option is not used, then matches are computed on the reverse complemented
strand.}

\Option{maxocc}{$\Showoptionarg{t}$}{
Specify the maximum number of occurrences of exact prefix matches (in case of
the matching statistics) or approximate prefix matches.
}

\Option{withwildcards}{}{
Output matches containing wildcard characters (e.g. N). This option cannot be
used in any of the following cases:
\begin{itemize}
\item
with option \Showoption{pck}, 
\item
in the \textit{ms}-mode,
\item
for all modes with \(k=0\).
\end{itemize}
}

\Option{best}{}{
For each tag, only show matches for the smallest possible distance. That is,
if a tag has exact matches in the input index, then only exact matches are
shown. If there are no exact matches, but matches with distance 1, then
only these are shown. If there are no matches matches with distance 1 (and
hence no exact matches), but with distance 2, then only these are shown etc.
}

\Option{output}{$\Showoptionarg{key}_{1}\ldots\Showoptionarg{key}_{q}$}{
Use combination of the following keywords to specify output according to
the following table:

\begin{center}
\begin{tabular}{l|l}
keyword & shows the following\\\hline
tagnum  &       show ordinal number of tag\\
tagseq   &      show tag sequence\\
dblength &      show length of match in database\\
dbstartpos&     show start position of match in database\\
abspos &        show absolute value of dbstartpos\\
dbsequence&     show sequence of match\\
strand    &     show strand\\
edist     &     show edit distance\\
tagstartpos&    show start position of match in tag (only for 
                \Showoption{maxocc})\\
taglength  &    show length of match in tag (only for option 
                \Showoption{maxocc})\\
tagsuffixseq&   show suffix tag involved in match (only for option
                \Showoption{maxocc})
\end{tabular}
\end{center}
This option only has an effect when used in the cdiff-mode. If in cdiff
mode this option is not used, then the output is such as if keywords
\texttt{tagnum}, \texttt{tagseq}, \texttt{dblength}, \texttt{dbstartpos},
\texttt{strand} were used.
}

\Helpoption

\end{Justshowoptions}
The following conditions must be satisfied:
\begin{enumerate}
\item
Option \Showoption{q} is mandatory.
\item
Either option \Showoption{pck} or \Showoption{esa} must be used. Both cannot
be combined.
\item
If option \Showoption{e} is not used, then option \Showoption{maxocc}
is required.
\end{enumerate}

\section{Examples}
Suppose that in some directory, say \texttt{homo-sapiens}, we have 24 gzipped
\Fasta files containing all 24 human chromomsomes. These may have been 
downloaded from
\url{ftp://ftp.ensembl.org/pub/current_fasta/homo_sapiens_47_36i/dna}.

In the first step, we construct the packed index for the entire human genome:

\begin{Output}
gt packedindex mkindex -dna -dir rev -parts 12 -bsize 10 -sprank -locfreq 32
                       -tis -ssp -indexname pck-human -db homo-sapiens/*.gz
\end{Output}

The program runs for little more than two hours and delivers 
an index \texttt{human-all} consisting of three files:

\begin{Output}
ls -lh human-all.*
-rw-r----- 1 kurtz gistaff   37 2008-07-13 17:00 pck-human.al1
-rw-r----- 1 kurtz gistaff 2.6G 2008-07-13 19:22 pck-human.bdx
-rw-r----- 1 kurtz gistaff 3.3K 2008-07-13 19:22 pck-human.prj
\end{Output}

Suppose that the compressed file \texttt{Q1.gz} contains short sequence tags
in multiple \Fasta format. In a first call we run \Program in \textit{ms}-mode:

\EXECUTE{gt tagerator -q Q1.gz -maxocc 10 -pck pck-human | head -n 25}

The first line shortly reports the kind of computation performed. The second
and third line give the name of the index containing the subject sequences.
It is reported whether it is a packed index or an enhanced suffix array. Then,
for each tag its length, say \(m\), and the tag \(p\) itself is shown, followed 
by a block of \(m\) lines each containing one integers. For all \(i\in[1,m]\), 
the first column in the \(i\)th line is the matching statistics length, say
\(l\). This means that \(\Substring{p}{i}{i+l-1}\) is the maximum length prefix
of \(\Substring{p}{i}{m}\) that occurs as a substring in the index. The
second column gives the strand of the match: a \texttt{+} stands for the
forward strand and a \texttt{-} for the reverse strand. If the number of 
occurrences of \(\Substring{p}{i}{i+l-1}\) is smaller than the maximum
occurrence parameter (\(10\) in the above case), then all matching statistics 
positions are reported in ascending order. If it is larger than the maximum
occurrence parameter, no positions are shown.
So, for example, the sequence \texttt{gcttgctg} of length 8 starting at tag
position 3 is the longest prefix of \texttt{gcttgctgctgca} that occurs as 
a substring in the index. It occurs at the positions 25583, 88695, 213546,
etc. Note that the matching statistics lines for the forward strand are
followed by the matching statistics block on the reverse strand. Note however,
that the matching statistics positions are always reported with respect to the
forward strand.

In a second call, we run \Program in \textit{cdiff}-mode:

\EXECUTE{gt tagerator -q Q1.gz -pck pck-human -e 2 | head -n 28}

The first line shortly reports the kind of computation performed. The second
and third line are as in the previous example. Then, for
each tag its length and the tag itself is shown, followed by 
a block of lines each containing three integers and one of the
symbols \texttt{+} or \texttt{-}, denoting the strand of the match.
Each such triple of integers reports the length of an approximate match,
the subject sequence in which the match occurs, and the relative position
of the match in this sequence. For each start position only the shortest 
length is reported. If there are no
approximate matches, then no such line appears. Note that the match always
refers to the complete pattern.

Our third example concerns the \textit{pdiff}-mode:

\EXECUTE{gt tagerator -q Q1.gz -pck pck-human -e 1 -maxocc 5 | head -n 20}

The output is similar as in the previous example, except that each match
is reported by four integers with a sign between the first three and the
last number. The first number reports the length of the
approximate match, the second reports the subject sequence number in
which the match occur, the third its relative position in the subject
sequence. The last number reports the length of the prefix 
of the tag involved in the approximate match. Note that in some cases,
the two length values are different.
\end{document}
