\documentclass[12pt]{article}
\usepackage[a4paper,top=20mm,bottom=20mm,left=20mm,right=20mm]{geometry}
\usepackage{url}
\usepackage{alltt}
\usepackage{xspace}
\usepackage{times}
\usepackage{listings}
\usepackage{bbm}
\usepackage{verbatim}
\usepackage{ifthen}
\usepackage{comment}
\usepackage{optionman}

\ifthenelse{\isundefined{\BuildMatstat}}{%
\includecomment{AboutUniquesub}
\excludecomment{AboutMatstat}
\newcommand{\AboutUniquesubcmd}[1]{#1}
\newcommand{\AboutMatstatcmd}[1]{}
}{%
\includecomment{AboutMatstat}
\excludecomment{AboutUniquesub}
\newcommand{\AboutMatstatcmd}[1]{#1}
\newcommand{\AboutUniquesubcmd}[1]{}
}

\newcommand{\Substring}[3]{#1[#2..#3]}

\begin{AboutUniquesub}
\newcommand{\Program}[0]{\texttt{uniquesub}\xspace}
\newcommand{\Mup}[1]{\mathit{mup(s,#1)}}
\title{\Program: a program for computing\\
       minimum unique substrings\\
       a manual}
\end{AboutUniquesub}

\begin{AboutMatstat}
\newcommand{\Program}[0]{\texttt{matstat}\xspace}
\newcommand{\MS}[1]{\mathit{ms(s,#1)}}
\title{\Program: a program for computing\\
       matching statistics\\
       a manual}
\end{AboutMatstat}

\author{\begin{tabular}{c}
         \textit{Stefan Kurtz}\\
         Center for Bioinformatics,\\
         University of Hamburg
        \end{tabular}}

\begin{document}
\maketitle

\section{The program \Program}

The program \Program is called as follows:
\par
\noindent\Program [\textit{options}] \Showoption{query} \Showoptionarg{files} [\textit{options}] 
\par
\Showoptionarg{files} is a white space separated list of at least one 
filename. Any sequence occurring in any file specified in \Showoptionarg{files}
is called \textit{unit} in the following.
In addition to the mandatory option \Showoption{query}, the program
must be called with either option \Showoption{pck} or \Showoption{esa}
which specify to use a packed index or an enhanced suffix array for 
a given set of subject sequences.

\begin{AboutUniquesub}
\Program computes for all positions \(i\) in each unit, say \(s\) of length
\(n\), the length \(\Mup{i}\) of the minimum unique prefix 
at position \(i\), if it exists. Uniqueness always refers to all substrings
represented by the index. \(\Mup{i}\) is defined by the following two 
statements:
\begin{itemize}
\item
If \(\Substring{s}{i}{n-1}\) is not unique in the index, then \(\Mup{i}=\bot\).
That is, it is undefined.
\item
If \(\Substring{s}{i}{n-1}\) is unique in the index, then \(\Mup{i}=m\), where 
\(m\) is the smallest value such that \(i+m-1\leq n-1\) and 
\(\Substring{s}{i}{i+m-1}\) occurs exactly once as a substring in the index.
\end{itemize}
Note that it is possible that for all \(i\in[0,n-1]\) we have 
\(\Mup{i}=\bot\), which means that unit \(s\) does not contain any unique 
substring. In this case, the program reports nothing for the corresponding
unit.
\end{AboutUniquesub}
\begin{AboutMatstat}
\Program computes the  \textit{matching statistics} for each unit. That is, 
for each position \(i\) in 
each unit, say \(s\) of length \(n\), \(\MS{i}=(l,j)\) is computed. Here
\(l\) is the largest integer such that \(\Substring{s}{i}{i+l-1}\) matches
a substring represented by the index and \(j\) is a start position of the
matching substring in the index. We say that \(l\) is the length of \(\MS{i}\)
and \(j\) is the subject position of \(\MS{i}\).
\end{AboutMatstat}

The following options are available in \Program:

\begin{Justshowoptions}
\begin{comment}
\Option{fmi}{$\Showoptionarg{indexname}$}{
Use the old implementation of the FMindex. This option is not recommended.
}
\end{comment}

\Option{esa}{$\Showoptionarg{indexname}$}{
Use the given enhanced suffix array to compute the matches.
}

\Option{pck}{$\Showoptionarg{indexname}$}{
Use the packed index (an efficient representation of the FMindex)
to compute the matches.
}


\Option{query}{$\Showoptionarg{files}$}{
Specify a white space separated list of query files containing the units.
At least one query file must be given. The files may be in 
gzipped format, in which case they have to end with the suffix \texttt{.gz}.
}

\begin{AboutUniquesub}
\Option{min}{$\ell$}{
Specify the minimum length $\ell$ of the minimum unique prefixes.
That is, for each unit \(s\) and each positions \(i\) in \(s\), the program 
reports the values \(i\) and \(\Mup{i}\) whenever \(\Mup{i}\geq\ell\). 
}

\Option{max}{$\ell$}{
Specify the maximum length $\ell$ of the minimum unique prefixes.
That is, for each unit \(s\) and each positions \(i\) in \(s\), the program 
reports the values \(i\) and \(\Mup{i}\) whenever \(\Mup{i}\leq\ell\).
}

\Option{output}{(\Showoptionkey{querypos}$\mid$\Showoptionkey{sequence})}{
Specify what to output. At least one of the two keys words
$\Showoptionkey{querypos}$ and $\Showoptionkey{sequence}$ must be used. 
Using the keyword $\Showoptionkey{querypos}$ shows the query position.
Using the keyword $\Showoptionkey{sequence}$ shows the sequence content
of the match.
}
\end{AboutUniquesub}

\begin{AboutMatstat}
\Option{min}{$\ell$}{
Specify the minimum value $\ell$ for the length of the matching statistics.
That is, for each unit \(s\) and each position \(i\) in \(s\), the program 
reports all values \(i\) and \(\MS{i}\) if the 
length of \(\MS{i}\) is at least \(\ell\).
}

\Option{max}{$\ell$}{
Specify the maximum length $\ell$ for the length of the matching statistics.
That is, for each unit \(s\) and each positions \(i\) in \(s\), the program 
reports the values \(i\) and \(\MS{i}\) if the length of \(\MS{i}\)
is at most \(\ell\).
}

\Option{output}{(\Showoptionkey{subjectpos}$\mid$\Showoptionkey{querypos}$\mid$\Showoptionkey{sequence})}{
Specify what to output. At least one of the three keys words
$\Showoptionkey{subjectpos}$,
$\Showoptionkey{querypos}$, and
$\Showoptionkey{sequence}$ must be used.
Using the keyword $\Showoptionkey{subjectpos}$ shows the 
subject position of the matching statistics.
Using the keyword $\Showoptionkey{querypos}$ shows the query position.
Using the keyword $\Showoptionkey{sequence}$ shows the sequence content
}
\end{AboutMatstat}

\Helpoption

\end{Justshowoptions}
The following conditions must be satisfied:
\begin{enumerate}
\item
Either option  \Showoption{min} or option \Showoption{max} must be used.
\item
If both options \Showoption{min} and \Showoption{max} are used, then
the value specified by option \(\Showoption{min}\) must be smaller
than the value specified by option \(\Showoption{max}\).
\item
Either option \Showoption{pck} or \Showoption{esa} must be used. Both cannot
be combined.
\end{enumerate}

\section{Examples}

Suppose that in some directory, say \texttt{homo-sapiens}, we have 25 gzipped
fasta files containing all 24 human chromomsomes plus one file with 
mitrochondrial sequences. These may have been downloaded from
\url{ftp://ftp.ensembl.org/pub/current_fasta/homo_sapiens_47_36i/dna}.

In the first step, we construct the packed index for the entire genome:

\begin{Output}
gt packedindex mkindex -dna -dir rev -parts 15 -bsize 10 -locfreq 32
                       -indexname human-all -db homo-sapiens/*.gz
\end{Output}

The program runs for almost two hours and delivers 
an index \texttt{human-all} consisting of three files:

\begin{Output}
ls -lh human-all.*
-rw-r----- 1 kurtz gistaff   37 2008-01-24 00:47 human-all.al1
-rw-r----- 1 kurtz gistaff 1.9G 2008-01-24 02:37 human-all.bdx
-rw-r----- 1 kurtz gistaff 3.4K 2008-01-24 02:37 human-all.prj
\end{Output}

This is used in the following call to the program \Program:

\begin{AboutUniquesub}
\begin{Output}
gt uniquesub -output querypos -min 20 -max 30 -query queryfile.fna 
             -pck human-all
unit 0 (Mus musculus, chr 1, complete sequence)
1007 20
1010 22
1011 22
1012 21
1013 21
...
\end{Output}

For all units \(s\) in the multiple \Fasta file \texttt{queryfile.fna},
a line is shown, reporting the number of the unit and the original fasta
header. Also, all for positions \(i\) in \(s\) satisfying 
\(20\leq \Mup{i}\leq 30\), \(i\) and \(\Mup{i}\) is reported.

The first column is the relative position in the unit sequence (counting
from 0). The second column shows the length value.

To additionally report the sequence content of the
minimum unique prefixes we add the keyword \Showoptionkey{sequence} to option
\Showoption{output}:

\begin{Output}
gt uniquesub -output querypos sequence -min 20 -max 30 
             -query queryfile.fna -pck human-all
unit 0 (Mus musculus, chr 1, complete sequence)
1007 20 ctgacagtttttttttttta
1010 22 acagttttttttttttacttta
1011 22 cagttttttttttttactttat
1012 21 agttttttttttttactttat
1013 21 gttttttttttttactttata
...
\end{Output}
\end{AboutUniquesub}

\begin{AboutMatstat}
\begin{Output}
gt matstat -output subjectpos querypos sequence -min 20 -max 30 
           -query queryfile.fna -pck human-all
unit 0 (Mus musculus, chr 1, complete sequence)
22 20 390765125 actgtatctcaaaatataaa
253 21 258488266 gggaataaacatgtcattgag
254 20 258488267 ggaataaacatgtcattgag
275 20 900483549 taattctatttttctttctt
480 20 1008274536 gcttgaagatcatgatccag
..
\end{Output}
Here, the first column shows the relative positions in unit 0 for which the
length of the matching statistics is between 20 and 30. The second column is
the corresponding length value. The third column shows position of the
matching sequence in the index, and the fourth shows the sequence content.
\end{AboutMatstat}
\end{document}
